% Section 3: Hoare's Notation and Basic Concepts
\section{Hoare's Notation}

\begin{frame}{Hoare's Notation}
    \begin{block}{Historical Context}
        C.A.R. Hoare introduced the following notation called a \textbf{partial correctness specification} for specifying what a program does:
        \begin{center}
            \Large $\hoare{P}{C}{Q}$
        \end{center}
    \end{block}
    
    \begin{block}{Components}
        \begin{itemize}
            \item $C$ is a command (a program or program fragment)
            \item $P$ and $Q$ are conditions on the program variables used in $C$
            \item $P$ is called the \textbf{precondition}
            \item $Q$ is called the \textbf{postcondition}
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}{The Precondition (P)}
    \begin{block}{Acceptable Initial State}
        The \textbf{precondition} defines the set of initial states for which the program is guaranteed to work correctly.
        \begin{itemize}
            \item It's an assumption about the values of program variables before execution.
            \item If the precondition is not met, the program has no obligations. It can crash, loop forever, or produce a wrong answer.
            \item Note: Reasoning about memory layout and heap requires \emph{Separation Logic}, an extension of Hoare Logic that can reason about pointer structures and memory allocation
        \end{itemize}
    \end{block}

    \begin{example}
        For a program that calculates the square root of X:
        \begin{center}
            Informal: ``X is greater than zero''\\
            Formal: $\pre{X > 0}$
        \end{center}
    \end{example}
\end{frame}

\begin{frame}{The Postcondition (Q)}
    \begin{block}{Acceptable Final State}
        The \textbf{postcondition} describes the state of the program after it has finished executing.
        \begin{itemize}
            \item It's the "promise" or "guarantee" of the specification.
            \item It typically relates the final values of variables to their initial values.
        \end{itemize}
    \end{block}

    \begin{example}
        For the square root program:
        \begin{center}
            Informal: ``Y is the square root of X''\\
            Formal: $\post{Y \times Y = X \wedge Y \ge 0}$
        \end{center}
        (Note: we relate the final value of Y to the initial value of X).
    \end{example}
\end{frame}

\begin{frame}{Writing Conditions}
    \begin{block}{Mathematical Notation}
        Conditions on program variables will be written using standard mathematical notations together with \textbf{logical operators}:
        \begin{itemize}
            \item $\wedge$ (and)
            \item $\vee$ (or)
            \item $\neg$ (not)
            \item $\Rightarrow$ (implies)
        \end{itemize}
    \end{block}
    
    \begin{example}
        Some example conditions:
        \begin{itemize}
            \item $x > 0 \wedge y \geq 0$ (x is positive AND y is non-negative)
            \item $x = 0 \vee y = 0$ (x equals zero OR y equals zero)
            \item $x > 0 \Rightarrow x^2 > 0$ (if x is positive, then x squared is positive)
        \end{itemize}
    \end{example}
\end{frame}

\begin{frame}{Formal Specification: The Hoare Triple}
    \begin{block}{Combining Pre- and Postconditions}
        Hoare Logic provides a formal notation to write specifications, called a \textbf{Hoare Triple}.
        \[ \hoare{P}{S}{Q} \]
        This is read as:
        \begin{quote}
            If the precondition $P$ is true before executing the program $S$, and if $S$ terminates, then the postcondition $Q$ will be true afterward.
        \end{quote}
    \end{block}

    \begin{example}[Square Root Specification]
        Combining our previous examples, the specification for a square root program $S$ is:
        \[ \hoare{X > 0}{S}{Y \times Y = X \wedge Y \ge 0} \]
        Here, $S$ is the placeholder for the actual program code (the "Action").
    \end{example}
\end{frame}

\begin{frame}{Evolution of Notation}
    \begin{block}{Historical Note}
        Hoare's original notation was $P\ \{C\}\ Q$ not $\hoare{P}{C}{Q}$, but the latter form is now more widely used.
    \end{block}
    
    \begin{block}{Alternative Notations}
        You may encounter different notations in the literature:
        \begin{itemize}
            \item Original: $P\ \{C\}\ Q$
            \item Modern: $\hoare{P}{C}{Q}$
            \item Some texts: $\{P\}\ C\ \{Q\}$ (without special formatting)
        \end{itemize}
        All represent the same concept: a partial correctness specification.
    \end{block}
\end{frame}

\begin{frame}{Partial Correctness}
    \begin{block}{What is Partial Correctness?}
        A Hoare triple $\hoare{P}{C}{Q}$ expresses \textbf{partial correctness}:
        \begin{quote}
            \emph{If} the precondition $P$ is true before executing command $C$, \emph{and if} $C$ terminates, \emph{then} the postcondition $Q$ will be true after execution.
        \end{quote}
    \end{block}
    
    \begin{alertblock}{Important: Termination Not Guaranteed}
        Partial correctness does \textbf{not} guarantee that the program terminates!
        \begin{itemize}
            \item It only says what must be true \emph{if} the program terminates
            \item A program that loops forever can still be partially correct
            \item Total correctness = Partial correctness + Termination
        \end{itemize}
    \end{alertblock}
\end{frame}

\begin{frame}{Reading Hoare Triples}
    \begin{block}{How to Read $\hoare{P}{C}{Q}$}
        The triple $\hoare{P}{C}{Q}$ can be read as:
        \begin{enumerate}
            \item ``If $P$ is true, then after $C$ executes, $Q$ will be true''
            \item ``$C$ transforms states satisfying $P$ into states satisfying $Q$''
            \item ``Starting from $P$, command $C$ establishes $Q$''
        \end{enumerate}
    \end{block}
    
    \begin{example}[Simple Assignment]
        $\hoare{x = 5}{\assign{y}{x + 1}}{y = 6}$
        
        This reads as: ``If $x$ equals 5 before the assignment, then $y$ will equal 6 after the assignment.''
    \end{example}
\end{frame}

\begin{frame}{Meaning of Hoare's Notation}
    \begin{block}{Formal Definition}
        $\hoare{P}{C}{Q}$ is true if:
        \begin{itemize}
            \item whenever $C$ is executed in a state satisfying $P$
            \item and \emph{if} the execution of $C$ terminates
            \item then the state in which $C$ terminates satisfies $Q$
        \end{itemize}
    \end{block}
    
    \begin{example}[Assignment Command]
        Consider: $\hoare{X = 1}{X := X + 1}{X = 2}$
        \begin{itemize}
            \item $P$ is the condition that the value of X is 1
            \item $Q$ is the condition that the value of X is 2
            \item $C$ is the assignment command $X := X + 1$ (i.e. `X becomes X+1')
        \end{itemize}
    \end{example}
\end{frame}

\begin{frame}{Truth and Falsity of Hoare Triples}
    \begin{example}[True Triple]
        $\hoare{X = 1}{X := X + 1}{X = 2}$ is \textbf{true}
        
        \textbf{Why?} Starting from a state where $X = 1$, executing $X := X + 1$ results in $X = 2$.
    \end{example}
    
    \begin{example}[False Triple]
        $\hoare{X = 1}{X := X + 1}{X = 3}$ is \textbf{false}
        
        \textbf{Why?} Starting from $X = 1$, executing $X := X + 1$ results in $X = 2$, not $X = 3$.
    \end{example}
    
    \begin{alertblock}{Key Insight}
        A Hoare triple is a mathematical statement that can be either true or false. It makes a claim about what happens when a program executes.
    \end{alertblock}
\end{frame}